---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# osfd

<!-- badges: start -->
<!-- badges: end -->

{**osfd**} (or: "oh yet another, spatial fisheries data package") consolidates
the ICES WGSFD VMS and logbook datacall workflow into a properly documented,
testable package. A junk drawer, finally organised into labelled boxes — and
then organised again, because the first attempt was also a junk drawer.

All exported functions carry an `fd_` prefix (fisheries data, or fishy data —
you decide). The package covers the full preprocessing and analysis pipeline:
QC and cleaning (Phase 2), trip linking, activity classification, landings
distribution, swept area, and effort enrichment (Phase 3).

## Installation

```r
# install.packages("pak")
pak::pak("einarhjorleifsson/osfd")
```

Some dependencies are not on CRAN and come from the ICES r-universe:

```r
install.packages(
  c("sfdSAR", "icesVMS", "icesVocab", "icesConnect"),
  repos = "https://ices-tools-prod.r-universe.dev"
)
```

## Quick start

The package ships with demo `eflalo` and `tacsat` datasets. A sketch of the
full pipeline:

```{r example, eval = FALSE}
library(osfd)
library(dplyr)
library(lubridate)

# --- Write demo data to disk (rds, one file per year) ------------------------
osfd::eflalo |>
  mutate(year = year(dmy(FT_DDAT))) |>
  fd_write_dataset("data-raw/eflalo", year_col = "year", name = "eflalo")

osfd::tacsat |>
  mutate(year = year(dmy(SI_DATE))) |>
  fd_write_dataset("data-raw/tacsat", year_col = "year", name = "tacsat")

# Parquet is also supported (requires the duckdbfs package):
# fd_write_dataset(..., format = "parquet")

# --- Preprocessing -----------------------------------------------------------
eflalo <- fd_read_dataset("data-raw/eflalo", years = 1803, name = "eflalo")
tacsat <- fd_read_dataset("data-raw/tacsat", years = 1803, name = "tacsat")

# Coerce columns and parse datetimes
eflalo <- fd_eflalo_clean(eflalo)
tacsat <- fd_tacsat_clean(tacsat)

# Single-pass QC (adds a `checks` column; caller decides what to filter)
eflalo <- fd_eflalo_checks(eflalo, iv_gear = NULL, iv_met6 = NULL, year = 1803)
eflalo |> count(checks)
eflalo <- filter(eflalo, checks == "10 no vessel tonnage") |> select(-checks)

# question here if checking interval is not premature
## Currently does not check if points on land (checked via ices_area)
tacsat <- fd_tacsat_checks(tacsat, it_min = 300)
tacsat |> count(checks)
tacsat <- filter(tacsat, checks == "ok") |> select(-checks, -INTV)

# --- Analysis ----------------------------------------------------------------
# Assign trip identifiers (and vessel attributes) to pings
tacsatp <- fd_assign_tripid(tacsat, eflalo)

# Calculate ping intervals and classify fishing activity
tacsatp <- fd_intv_tacsat(tacsatp, level = "trip", fill.na = TRUE)

# TODO: Set the ping states (fishing or not)
# ...

# Distribute landings among fishing pings
tacsatEflalo <- fd_split_pings(tacsatp, eflalo)

# Gear width and swept area
tacsatEflalo$GEARWIDTHKM <- fd_gearwidth(tacsatEflalo)
tacsatEflalo$SA_KM2 <- fd_swept_area(
  gear          = tacsatEflalo$LE_GEAR,
  fishing_hours = tacsatEflalo$INTV / 60,
  gear_width    = tacsatEflalo$GEARWIDTHKM,
  fishing_speed = tacsatEflalo$SI_SP
)
```

## Datacall flow

This preliminarily placed here, just a bookkeeping of the current datacall flow with comments if not yet implemented in {osfd}

### Preprocessing

#### tacsat

* 1.2.1 Remove VMS pings outside the ICES areas
  * This can be an expensive process if ais data is rich, because dataframe turned to sf, then a join with ices-area shapefile but then geometry is droppped. Question for now is if this can be moved more downstream, where other spatial acrobatics take place.
  * NOTE: Not yet implemented in {osfd}
* 1.2.2 Remove duplicate records - in osfd::fd_tacsat_check
* 1.2.3 Remove points that have impossible coordinates
  * this is a redundant step, given 1.2.1 above
* 1.2.4 Remove points which are pseudo duplicates as they have an interval rate < x minutes - in osdf:fd_tacsat_check
  * the check function has accepts the minimum interval
* 1.2.5 Remove points in harbour
  



## Source

Functions are ported from
[ICES-VMS-and-Logbook-Data-Call](https://github.com/ices-eg/ICES-VMS-and-Logbook-Data-Call)
scripts `0_global.R`, `1_eflalo_tacsat_preprocessing.R`, and
`2_eflalo_tacsat_analysis.R`. Modernised script versions using `osfd` functions
are in `inst/scripts/`.

## License

CC BY 4.0
