---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r knitr, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

# osfd

<!-- badges: start -->
<!-- badges: end -->

{**osfd**} (or: "oh yet another, spatial fisheries data package") consolidates
the ICES WGSFD VMS and logbook datacall workflow into a properly documented,
testable package. A junk drawer, finally organised into labelled boxes — and
then organised again, because the first attempt was also a junk drawer.

All exported functions carry an `fd_` prefix (fisheries data, or fishy data —
you decide). The package covers the QC preprocessing and analysis pipeline of
the ICES VMS/logbooks data-call:
QC and cleaning (Phase 1), trip linking, activity classification, landings
distribution, swept area, and effort and catch enrichment (Phase 2).

The process here will definitively **not reproduce** to the full extent the
official [data-call process](https://github.com/ices-eg/ICES-VMS-and-Logbook-Data-Call).

## Installation

Some dependencies are not on CRAN and come from the ICES r-universe:

```r
... instructions pending for non-crans
# install.packages("remotes")
remotes::install_github("einarhjorleifsson/osfd")
```

## Setting the stage

Here load the needed libraries and create some temporary demo data. We are going
to store them in directory "_garbage", set that up manually a priori.

```{r load}
# library(conflicted).  # don't ever use this again
library(sf)
library(nanoparquet)
library(tidyverse)
library(osfd)
library(vmstools)
```

```{r overwrite_default_data}
harbours_nw <- 
  read_sf("~/Documents/stasi/fishydata/data/ports/ports_iceland_faroe.gpkg")
```
## Preprocessing

Here we work with Icelandic dataset that has been painfully converted to the EFLALO2 and TACSAT2 format.

### Data setup

```{r setup}
eflalo <- "data-raw/eflalo_IS.parquet" |>  
  read_parquet() |> 
  # osfd::eflalo |> 
  fd_clean_eflalo()
tacsat <- 
  "data-raw/tacsat_IS.parquet" |> 
  read_parquet() |> 
  #osfd::tacsat |> 
  fd_clean_tacsat()

# Tidy alternative
trips <- eflalo |> 
  fd_trips()
events <- eflalo |> 
  fd_events()  
```

### Data checks

```{r checks}
# Classical
eflalo_classic <- eflalo |> 
  fd_flag_eflalo()
eflalo_classic |> count(.checks, name = "records") |> knitr::kable(caption = "Eflalo QC: Data checks")
# Tidy
trips <- trips |> 
  fd_flag_trips()
trips |> count(.tchecks, name = "trips") |> knitr::kable(caption = "Trips QC: Data checks")
events <- events |> 
  fd_flag_events()
events |> count(.echecks, name = "events") |> knitr::kable(caption = "Events QC: Data checks")

# the good boy
tacsat <- tacsat |> 
  fd_flag_tacsat()
tacsat |> count(.checks, name = "pings") |> knitr::kable(caption = "Trail QC: Data checks")
```

### Filtering

* Objective: Pass data to the next step (analysis)
* Two alternative (I am "thinking" which would be less painful, coding-wise):
  * Use the classical approach
  * Use the tidy approach
  
```{r filter}
# Classical
eflalo_dropped <- eflalo_classic |> 
  filter(.checks != "ok")
eflalo_classic <- eflalo_classic |> 
  filter(.checks == "ok") |> 
  select(-.checks)
# Side-step: Possibly for diagnostics/comparison between classical/tidy
eflalo_alt <-     # name it this rather than tidy - because once joined the table is again non-tidy
  trips |> 
  filter(.tchecks == "ok") |> 
  inner_join(events |> 
              filter(.echecks == "ok"))
# Tidy
trips <- trips |> 
  filter(.tchecks == "ok") |> 
  select(-.tchecks)
events <- events |> 
  filter(.echecks == "ok") |> 
  select(-.echecks)

# same as usual
tacsat <- tacsat |> 
  filter(.checks == "ok") |> 
  select(-.checks)
  
```

## Analysis

... pending, but want to get Preprocessing right from start (although that may have to be revisited

## License

CC BY 4.0
