---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r knitr, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

# osfd

<!-- badges: start -->
<!-- badges: end -->

{**osfd**} (or: "oh yet another, spatial fisheries data package") consolidates
the ICES WGSFD VMS and logbook datacall workflow into a properly documented,
testable package.

All exported functions carry an `fd_` prefix (fisheries data, or fishy data —
you decide). The package current versions covers the QC pre-processing of
the ICES VMS/logbooks data-call. Processing functions are pending.

The process here will definitively **not reproduce** to the full extent the
official [data-call process](https://github.com/ices-eg/ICES-VMS-and-Logbook-Data-Call).

## Installation

Some dependencies are not on CRAN and come from the ICES r-universe:

```r
... instructions pending for non-crans
# install.packages("remotes")
remotes::install_github("einarhjorleifsson/osfd")
```

## Setting the stage

Here load the needed libraries and load data.

```{r load}
library(sf)
library(arrow)
library(tidyverse)
library(osfd)
```

```{r shapes}
ports  <- "https://heima.hafro.is/~einarhj/data/ports_iceland_faroe.gpkg" |> 
  read_sf() |> 
  select(port = pid)
areas <- read_sf("data-raw/ices_areas.gpkg") |> 
  select(area = Area_27)
depth <- read_sf("data-raw/gebco_ices.gpkg")
eusm  <- read_sf("data-raw/eusm.gpkg") |> # criminal size: ~120 million points!
  filter(MSFD_BBHT != "")
```

```{r data}
eflalo       <- "https://heima.hafro.is/~einarhj/data/eflalo_IS.parquet" |> 
  arrow::read_parquet()
tacsat       <- "https://heima.hafro.is/~einarhj/data/tacsat_IS.parquet" |> 
  arrow::read_parquet()
# fishing speed by gear and target
state_lookup <- "https://heima.hafro.is/~einarhj/data/gear_mapping.parquet" |> 
  arrow::read_parquet() |> 
  select(gear, target, s1, s2)
benthis_lookup <- fd_benthis_lookup()
```

## Pre-processing

... text is pending, but here is an illustration of user code flow to complete the task.

```{r preprocessing}
ais <- tacsat |> 
  fd_clean_tacsat() |> 
  fd_flag_tacsat(no_hands = TRUE,
                 minimum_interval_seconds = 30, 
                 areas = areas, ports = ports)

eflalo_clean <- eflalo |> fd_clean_eflalo()

trips <- eflalo_clean |> 
  fd_trips() |> 
  fd_flag_trips(no_hands = TRUE)

events <- eflalo_clean |>
  fd_events() |> 
  fd_flag_events(
    no_hands = TRUE,
    gear = icesVocab::getCodeList("GearType")$Key,
    met6 = icesVocab::getCodeList("Metier6_FishingActivity")$Key)
```

## Processing

A little peek-a-boo, ... not run, only partially ready

```{r, eval = FALSE}
ais2 <- ais |> 
  fd_add_trips(trips, cn = c("tid", "length", "kw", "gt", ".tid")) |> 
  # some monkey-buissness - run fd_check_events_join(ais2, events) to see issues
  fd_add_events(events, resolve = TRUE) |> 
  # here the second intv call is made
  group_by(.tid) |> 
  mutate(dt_sec = fd_interval_seconds(time),
         kwh <- kw * dt_sec / 60^2) |> 
  ungroup() |> 
  fd_add_state(state_lookup) |> 
  filter(state == "fishing") |> 
  # fd_add_catch(events) |>                # pending, resolve fd_add_events first
  fd_add_sf(eusm) |> 
  fd_add_sf(depth) |> 
  mutate(csq = fd_calc_csq(lon, lat)) #|>
# fd_add_gearwith(gear_width_table) |>   # pending
# fd_calc_sa()                           # pending
```

## Submission

More peek-a-boo, ... not run, not ready

```{r, eval = FALSE}
# ... pending
ais3 <- ais2 |> 
  fd_final_tests() 
ais3 |> fd_aggregate() |> fd_export_table1()
ais3 |> fd_aggregate() |> fd_export_table2()
```


## Some testing

```{r}
# sidestep in order to run datacall way below
e_test <- left_join(trips, events) |> 
  fd_revert_eflalo() |> 
  as.data.frame()
t_test <- ais |> 
  fd_revert_tacsat() |> 
  as.data.frame()

# Datacall way -----------------------------------------------------------------
merged <- vmstools::mergeEflalo2Tacsat(e_test, t_test)
cols <- c("LE_GEAR", "LE_MSZ", "VE_LEN", "VE_KW", "LE_RECT", "LE_MET", "LE_WIDTH", "VE_FLT", "VE_COU")
# Use a loop to add each column
for (col in cols) {
  # Match 'FT_REF' values in 'tacsatp' and 'eflalo' and use these to add the column from 'eflalo' to 'tacsatp'
  merged[[col]] <- e_test[[col]][match(merged$FT_REF, e_test$FT_REF)]
}

# {osfd} way - could be wrapped in a function ----------------------------------
aism <- ais |> 
  left_join(trips,
            by = join_by(cid, vid,
                         between(time, T1, T2)),
            relationship = "many-to-one")
# similarly, wrapped in a function
ais_trips <- ais |> fd_add_trips(trips, cn = c("tid", "length", "kw", "gt", ".tid"))

# comparison -------------------------------------------------------------------
identical(merged$FT_REF, replace_na(aism$tid, "0"))
```

## Small print

```{r}
devtools::session_info()
```


## License

CC BY 4.0
