---
title: "The ICESrectangle zone-A bug"
vignette: >
  %\VignetteIndexEntry{The ICESrectangle zone-A bug}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
```

## Background

ICES statistical rectangles are the standard spatial grid for European fisheries
data, including the VMS/logbook data call. Every VMS ping and every logbook event
is assigned an ICES rectangle code such as `"41E9"` or `"57A2"`.

The rectangle code has four parts:

| Part | Example | Meaning |
|------|---------|---------|
| Two-digit number | `57` | Latitude band (0.5° increments, starting at 36°N) |
| Letter | `A`–`M` (no `I`) | Longitude zone (nominally 10° wide) |
| Single digit | `0`–`9` | 1° column within the zone |

The key word is *nominally*. Zone A is a special case.

## Zone A is only 4 degrees wide

The western boundary of the ICES statistical area is **44°W**, not 50°W. Zone A
therefore spans only **-44° to -40°** — four degrees, not ten. Consequently,
zone A contains only **four** valid rectangles per latitude band (A0–A3), each
representing a 1° column:

| Rectangle | Longitude range |
|-----------|-----------------|
| A0        | -44° to -43°    |
| A1        | -43° to -42°    |
| A2        | -42° to -41°    |
| A3        | -41° to -40°    |

Codes A4 through A9 do not exist.

## The bug in `ICESrectangle()`

The `ICESrectangle()` function from `vmstools` (widely used in the ICES data
call processing chain) incorrectly extends zone A all the way to **-50°W** and
applies the same 10° modular arithmetic it uses for every other zone. For the
actual ICES area (within -44° to -40°), this produces digit offsets of 6–9
instead of the correct 0–3.

```{r, echo = FALSE}
library(dplyr)
library(osfd)

ICESrectangle_orig <- function(dF) {
  rectChar1n2 <- sprintf("%02i", as.integer(2 * (dF[, "SI_LATI"] - 35.5)))
  rectChar3 <- ifelse(dF[, "SI_LONG"] > -50 & dF[, "SI_LONG"] <= -40, "A",
    ifelse(dF[, "SI_LONG"] > -40 & dF[, "SI_LONG"] <= -30, "B",
      ifelse(dF[, "SI_LONG"] > -30 & dF[, "SI_LONG"] <= -20, "C",
        ifelse(dF[, "SI_LONG"] > -20 & dF[, "SI_LONG"] <= -10, "D",
          ifelse(dF[, "SI_LONG"] > -10 & dF[, "SI_LONG"] < 0, "E",
            ifelse(dF[, "SI_LONG"] >= 0 & dF[, "SI_LONG"] < 10, "F",
              ifelse(dF[, "SI_LONG"] >= 10 & dF[, "SI_LONG"] < 20, "G",
                ifelse(dF[, "SI_LONG"] >= 20 & dF[, "SI_LONG"] < 30, "H", "J"))))))))
  rectChar4 <- rep(NA, nrow(dF))
  idxlowzero  <- which(dF[, "SI_LONG"] < 0)
  idxabozero  <- which(dF[, "SI_LONG"] >= 0)
  if (length(idxlowzero) > 0)
    rectChar4[idxlowzero] <- ceiling(dF[idxlowzero, "SI_LONG"] %% 10 - 1 + 10) %% 10
  if (length(idxabozero) > 0)
    rectChar4[idxabozero] <- floor(dF[idxabozero, "SI_LONG"] %% 10)
  paste(rectChar1n2, rectChar3, rectChar4, sep = "")
}
```

```{r}
# The four 1-degree columns that actually exist in zone A (midpoints shown)
tibble(
  lon_midpoint = seq(-43.5, -40.5, by = 1),
  lat = 64
) |>
  mutate(
    ICESrectangle = ICESrectangle_orig(
      data.frame(SI_LONG = lon_midpoint, SI_LATI = lat)
    ),
    d2ir = d2ir(lon_midpoint, lat)
  )
```

Every zone-A record processed through `ICESrectangle()` carries an impossible
rectangle code. The digit is wrong by exactly 6 (the offset between where the
function *thinks* zone A starts, at -50°, and where it actually starts, at -44°).

## A secondary issue: the zone boundary convention

There is a second, smaller discrepancy at the boundaries between zones. The
legacy function uses `> lower & <= upper` for negative-longitude zones and
`>= lower & < upper` for non-negative zones. This inconsistency means a point
at exactly **-40°**, **-30°**, **-20°**, or **-10°** is placed in the *western*
zone, while `d2ir` (following `geo::d2ir`) places it in the *eastern* zone via
a `+1e-6` nudge.

```{r}
tibble(lon = c(-40.001, -40.000, -39.999), lat = 62) |>
  mutate(
    ICESrectangle = ICESrectangle_orig(
      data.frame(SI_LONG = lon, SI_LATI = lat)
    ),
    d2ir = d2ir(lon, lat)
  )
```

In practice, pings rarely land on exact integer longitudes, so this matters less
than the zone-A digit error. It is worth noting nonetheless when comparing
datasets processed by different tools.

## The corrected function: `d2ir()`

`osfd` provides `d2ir(lon, lat)` as a drop-in replacement. It follows the
`geo::d2ir` convention:

- Zone A uses **4° sub-divisions** (digits 0–3), matching the actual ICES extent.
- Boundary points receive the `+1e-6` nudge and fall into the eastern rectangle.
- Positions outside the ICES area (lat < 36, lat ≥ 85.5, lon ≤ -44, lon > 68.5)
  return `NA` rather than a nonsensical code.
- All operations (`floor()`, `%%`, `case_when()`, `paste0()`) are SQL-translatable,
  so the function works inside `dplyr::mutate()` on both in-memory data frames and
  lazy DuckDB tables.

```{r}
tibble(
  lon = c(-43.5, -5.0, 10.0, 25.0),
  lat = c( 64.0, 63.2, 58.0, 60.0)
) |>
  mutate(rect = d2ir(lon, lat))
```

## Implication for historical data call outputs

Any catch or effort aggregated by `LE_RECT` using `ICESrectangle()` in the
zone-A area will have been attributed to non-existent rectangle codes. Depending
on how downstream products consume the rectangle field — joining against ICES
reference grids, plotting on maps, or aggregating to sub-area level — these
records may have been silently dropped or miscounted.

The affected geographic area is small (roughly the southern tip of Greenland
and the waters around Iceland's east coast extending to the Faroe–Iceland
ridge), but it covers an important fishing region for several ICES member states.
Re-running the data call pipeline with `d2ir()` in place of `ICESrectangle()`
will produce corrected rectangle codes for all zone-A positions.
```
