---
title: "osfd vs. the official ICES datacall scripts"
vignette: >
  %\VignetteIndexEntry{osfd vs. the official ICES datacall scripts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  eval      = FALSE   # most code blocks are illustration only
)
```

## Overview

The official ICES VMS and Logbook Data Call workflow
(<https://github.com/ices-eg/ICES-VMS-and-Logbook-Data-Call>) is distributed as
four standalone R scripts (`0_global.R`, `1_eflalo_tacsat_preprocessing.R`,
`2_eflalo_tacsat_analysis.R`, `3_data_submission.R`) that analysts copy and
adapt for their national datasets.

**osfd** pursues the same end goal — producing spatially explicit effort and
catch distributions from VMS and logbook data — but takes a different
architectural path: a documented R package with consistent function interfaces,
leaving the analyst's script as a thin workflow driver rather than a
customisation target.

### Code size

The table below summarises non-blank, non-comment lines in each codebase.
"Roxygen" is counted separately because those lines are
documentation, not executable code.

```{r loc-table, eval = TRUE, echo = FALSE}
loc <- data.frame(
  Source = c(
    "osfd `R/`",
    "ICES datacall scripts"
  ),
  `Total lines` = c(2525, 3382),
  `Blank`       = c(181,  724),
  Roxygen       = c(1465, 224),
  Comments      = c(82,   497),
  `Code lines`  = c(797,  1937),
  check.names   = FALSE
)
knitr::kable(loc)
```

The ICES scripts contain roughly **2.4× more executable code** than the osfd
package source. Most of the overhead comes from inline QC tracking
(`remrecsTacsat`, `remrecsEflalo` matrices), year-loop boilerplate, file I/O,
and the activity-detection block. The osfd package instead moves that logic
inside documented functions, so the analyst's script stays short — the
pre-processing stage in `README.Rmd` fits in about 15 lines.

---

## Key differences

### 1 — Package vs. scripts

The datacall scripts are designed to be **copied and edited**. Each analyst
embeds their national configuration (threshold values, paths, gear lists) in
`0_global.R` and modifies the downstream scripts as needed. This is
practical for a multi-country submission process but makes upgrading to a
new version of the workflow a manual diff exercise.

osfd is an **R package**. Configuration is passed as function arguments; the
analyst's script calls documented functions with known inputs and outputs.
Upgrading means `remotes::install_github()`.

### 2 — Data model: monolithic eflalo vs. trips + events

The datacall scripts keep eflalo as a single, wide data frame throughout.
Vessel/trip-level attributes (departure time, arrival time, vessel length, engine
power) sit in the same row as gear/rectangle/date/catch columns, and the
same row is duplicated for every gear × date × rectangle combination within a
trip. All QC is applied to this monolithic structure.

osfd **decomposes** eflalo at the start:

```{r decompose}
eflalo_clean <- eflalo |> fd_clean_eflalo()

trips  <- eflalo_clean |> fd_trips()   # one row per trip
events <- eflalo_clean |> fd_events()  # one row per gear × date × rectangle
```

This makes trip-level checks (overlapping departure/arrival windows) and
event-level checks (gear validity, time-within-trip) independent of each
other. `fd_flag_eflalo()` orchestrates both and then applies the two
cross-level checks that need the full structure (catch date outside trip
window).

### 3 — Column naming

The datacall scripts use ICES ALLCAPS column names throughout
(`SI_LATI`, `FT_DDATIM`, `LE_GEAR`, …). osfd translates these to short
lowercase names immediately after loading:

| Raw ICES name | osfd name |
|---|---|
| `VE_REF` | `vid` |
| `SI_LATI` / `SI_LONG` | `lat` / `lon` |
| `FT_DDATIM` / `FT_LDATIM` | `T1` / `T2` |
| `LE_GEAR` | `gear` |
| `LE_RECT` | `ir` |

The original names are restored by `fd_revert_tacsat()` / `fd_revert_eflalo()`
for anything that still expects the legacy format (e.g. `vmstools::mergeEflalo2Tacsat()`).

### 4 — QC flagging: inline filtering vs. diagnostic/production modes

The datacall scripts filter records immediately when a check fails, then
record how many rows were removed in the `remrecsTacsat` / `remrecsEflalo`
matrices:

```r
# Datacall pattern — check and remove in one step
tacsat <- tacsat[lengths(overs) > 0, ]
remrecsTacsat["outsideICESarea", ] <- c(nrow(tacsat), percentage_remaining)
```

osfd separates **inspection** from **filtering** through the `no_hands`
parameter available on all four `fd_flag_*` functions:

```{r no-hands}
# Inspect failures before deciding what to remove
ais <- tacsat |>
  fd_clean_tacsat() |>
  fd_flag_tacsat(areas = areas, ports = ports, no_hands = FALSE)

ais |> count(.checks, sort = TRUE)

# Then filter manually — or set no_hands = TRUE to filter automatically
ais <- filter(ais, .checks == "ok") |> select(-.checks)
```

The `.checks` column carries a self-describing label (`"01 outside ICES area"`,
`"04 harbour ping"`, …) rather than a row count. This makes it easier to
examine failures in context before discarding records.

### 5 — VMS–logbook linking

The datacall scripts use `vmstools::mergeEflalo2Tacsat()` to assign a trip
identifier to each VMS ping. The function loops over vessels, binary-searches
the sorted trip table per vessel, then does a second loop and a `pmatch()` for
within-trip attribute assignment, followed by `trip_assign()` for pings
that fall on multi-gear or multi-rectangle trips. The full merge + enrichment
block runs to roughly 60 lines in `2_eflalo_tacsat_analysis.R`.

osfd uses dplyr's native **interval join** throughout:

```{r add-trips}
ais2 <- ais |>
  fd_add_trips(trips, cn = c("tid", "length", "kw", "gt", ".tid")) |>
  fd_add_events(events)
```

`fd_add_trips()` uses `join_by(cid, vid, between(time, T1, T2))` to assign
the trip identifier in a single pass. `fd_add_events()` then links each ping to
its logbook event using the more precise `between(time, t1, t2)` window derived
from `LE_STIME`/`LE_ETIME`, or a day-wide dummy window when those fields are
absent. Both functions include row-count guards that catch overlapping trip
windows or conflicting event windows before they silently inflate the data.

### 6 — Activity classification

The datacall scripts use a **mixture-model** approach (`mixtools`-backed
`ac.tac.anal()` / `act.tac()` from vmstools) for gears listed in
`autoDetectionGears`, and a speed-threshold lookup for the remainder. The
mixture model fits a bimodal (or tri-modal) Gaussian to the vessel speed
distribution to locate the fishing and steaming peaks automatically, with
analyst visual inspection to confirm the result. This is sophisticated but
requires `vmstools`, `mixtools`, and a manual review step.

osfd uses a simple speed-threshold join via `fd_add_state()`:

```{r state}
state_lookup <- read_parquet("gear_mapping.parquet") |>
  select(gear, target, s1, s2)

ais2 <- ais2 |>
  fd_add_state(state_lookup) |>
  filter(state == "fishing")
```

The lookup table is expected to be pre-built — by the analyst, using the
mixture model or any other method — and supplied as a plain data frame. The
separation keeps the classification logic out of the package and lets the
analyst use whatever method they trust.

### 7 — Catch distribution

The datacall scripts use `vmstools::splitAmongPings()` to distribute logbook
catch proportionally across fishing pings, dispatching at the most fine-grained
available level (trip × day × ICES rectangle → trip → vessel → fleet) with a
`conserve` option that retains catches from trips without VMS coverage.

The osfd equivalent (`fd_split_among_pings()`) is the primary outstanding work
item. The design is documented in `_articles/catch-distribution.Rmd`; the
interface will use the `trips`/`events` decomposition described above rather
than the monolithic eflalo data frame.

---

## Dependencies

```{r deps-table, eval = TRUE, echo = FALSE}
deps <- data.frame(
  Package     = c("dplyr / tidyr / lubridate", "sf", "data.table", "vmstools",
                  "doBy", "mixtools", "sfdSAR", "icesVMS", "icesVocab"),
  `datacall`  = c("✓", "✓", "✓", "✓ (core)", "✓", "✓", "✓", "✓", "✓"),
  `osfd`      = c("✓ (core imports)", "✓ (core import)", "—", "— (optional via revert)",
                  "—", "—", "✓ (suggested)", "✓ (suggested)", "✓ (suggested)"),
  check.names = FALSE
)
knitr::kable(deps)
```

The datacall workflow requires `vmstools` for `mergeEflalo2Tacsat()`,
`splitAmongPings()`, `sfsortTacsat()`, `intervalTacsat()`, and the activity
detection functions. `data.table` is used extensively for the overlap-trip and
catch-split logic. osfd drops both hard dependencies: the interval join replaces
`mergeEflalo2Tacsat()`, and `fd_calc_csq()` replaces `vmstools::CSquare()`
with a dbplyr-compatible implementation.

---

## What the analyst sees

The table below compares the user-facing code for the pre-processing stage
(the most developed part of osfd at this point).

**Datacall** (`1_eflalo_tacsat_preprocessing.R`, simplified):

```r
# Year loop
for (year in yearsToSubmit) {
  load(paste0(dataPath, "tacsat_", year, ".RData"))
  load(paste0(dataPath, "eflalo_", year, ".RData"))
  tacsat <- formatTacsat(tacsat)
  eflalo <- formatEflalo(eflalo)

  # TACSAT checks — inline filter + matrix tracking
  remrecsTacsat["total", ] <- c(nrow(tacsat), "100%")
  overs  <- sf::st_intersects(tacsat_sf, ices_areas_sf)
  tacsat <- tacsat[lengths(overs) > 0, ]
  remrecsTacsat["outsideICESarea", ] <- c(nrow(tacsat), ...)
  tacsat$unique_id <- paste(tacsat$VE_REF, ...)
  tacsat <- tacsat[!duplicated(tacsat$unique_id), ]
  # ... ~150 more lines for tacsat + eflalo checks ...
  save(tacsat, file = paste0(outPath, "cleanTacsat", year, ".RData"))
  save(eflalo, file = paste0(outPath, "cleanEflalo", year, ".RData"))
}
```

**osfd** (`README.Rmd`):

```r
ais <- tacsat |>
  fd_clean_tacsat() |>
  fd_flag_tacsat(areas = areas, ports = ports, no_hands = TRUE)

eflalo_clean <- eflalo |> fd_clean_eflalo()

trips <- eflalo_clean |>
  fd_trips() |>
  fd_flag_trips(no_hands = TRUE)

events <- eflalo_clean |>
  fd_events() |>
  fd_flag_events(no_hands = TRUE,
                 gear = icesVocab::getCodeList("GearType")$Key,
                 met6  = icesVocab::getCodeList("Metier6_FishingActivity")$Key)
```

The osfd version processes all years in a single pass (assuming the data are
read in together), requires no intermediate file I/O, and expresses the full
pre-processing in 15 lines. The datacall version offers more granular per-year
audit trails and is closer to the format required for national submission
reports; those are valid trade-offs depending on the analyst's needs.
