---
title: "Do we still need c-squares?"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## The question

The ICES VMS and Logbook Data Call requires catch and effort to be submitted
as gridded summaries, with each grid cell identified by a c-square code at
0.05° resolution — a 14-character string such as `"7500:104:100:1"`. This has
been the format since the data call was established, and it has been used
without much question ever since.

It is worth asking whether this is still the right choice, or whether it is
simply the choice that was made and never revisited.

This document argues that, for all intermediate analytical steps of the VMS
workflow, c-squares should be replaced by a simpler numeric midpoint encoding.
C-square codes should be generated only at the final submission step — as a
thin formatting layer applied to an already-aggregated result — if and when
the data call continues to require them. The case rests on four points: the
tool was designed for something else entirely, its boundary behaviour is
asymmetric, implementations disagree, and it is dramatically slower than
the alternative. A fifth point, addressed at the end, is that the data call
is heading toward a world in which the performance argument becomes not merely
inconvenient but untenable.

## Square peg, round hole: what c-squares were designed for

C-squares were proposed by Rees (2003) as a **metadata** spatial indexing
system — a compact text-based way to describe the geographic footprint of an
oceanographic dataset, so that catalogue records could be searched with nothing
more than a string prefix query. The motivating examples in the original paper
are indexing bird sightings and National Park extents for catalogue-level
discovery. It is therefore something of a surprise to find the format applied
to tens of millions of individual VMS pings per year.

The specification is candid about the intended scale. It recommends "a
resolution of at least 1 × 1 degree squares" for creators wishing to guarantee
interoperability with distributed c-squares search. Finer resolutions are
permitted, but the spec flags them as a resource concern: cell counts increase
by ×4 or ×25 with each decrease in resolution, "either requiring an equivalent
increase in computing resources or possibly slower addressing times."

The ICES VMS data call uses **0.05°** — four levels deep in the c-squares
hierarchy, and 400 times finer than the recommended interoperability floor.
This is a subtlety that appears not to have received much attention when the
data call format was chosen.

## The boundary problem

C-squares encode coordinates using a hierarchical scheme built around the
global quadrant system (NE, SE, SW, NW). This quadrant structure forces an
asymmetric boundary convention. The specification (v1.1, §9.3) states:

> Points "on the line" are normally encoded within the next "higher" square,
> i.e. further away from the global origin. A point at +10 latitude will be
> encoded within the ten degree square covering +10 to +20, not 0 to +10; a
> point at −10 latitude will be encoded within the ten degree square covering
> −10 to −20, not 0 to −10.

In plain terms: boundary points go to the cell with the **larger absolute
coordinate value**, regardless of sign. The consequence is that cell intervals
are left-closed in the positive half-plane and right-closed in the negative
half-plane:

- `[0.00, 0.05)`, `[0.05, 0.10)`, … (positive)
- `(−0.10, −0.05]`, `(−0.05, 0.00]`, … (negative)

This is different from a standard raster or `cut()`, which uses left-closed
intervals everywhere. Two positions separated by 0.000001° — one exactly on a
boundary, one just inside the next cell — land in different c-squares:

```{r setup}
library(osfd)
library(csquares)
library(tidyverse)
library(ggmisc)  #pak::pak("einarhjorleifsson/ggmisc")
```

```{r boundary-demo}
lons <- c(-0.05, -0.049999)
lats <- c(-0.005, -0.005)

fd_calc_csq(lons, lats, degrees = 0.05)
```

Plotted on a grid, the asymmetry is immediately visible. Each point below is
coloured by its c-square code; grid lines are drawn at multiples of 0.05°.
Points at `lon = −0.05` belong to the cell whose centre is at `−0.075`, not
`−0.025`. The grid lines do not split the cells they appear to delineate,
because the boundary convention flips direction at zero.

```{r asymmetric-plot, fig.cap = "Points coloured by c-square code near the origin. Grid lines mark cell boundaries (multiples of 0.05°). Because the boundary convention flips at zero, grid lines do not bisect the coloured bands they appear to separate."}
dx <- dy <- 0.05
x1 <- y1 <- -0.10
x2 <- y2 <-  0.10

g <-
  expand_grid(
    lon = seq(x1, x2, by = 0.01),
    lat = seq(y1, y2, by = 0.01)
  ) |>
  mutate(cs = fd_calc_csq(lon, lat, dx))

ggplot() +
  theme_grey() +
  geom_vline(xintercept = seq(x1, x2, by = dx), colour = "grey60") +
  geom_hline(yintercept = seq(y1, y2, by = dy), colour = "grey60") +
  geom_point(data = g, aes(lon, lat, colour = cs), size = 3.5) +
  ggmisc::scale_colour_crayola() +
  scale_x_continuous(breaks = seq(x1, x2, by = dx)) +
  scale_y_continuous(breaks = seq(y1, y2, by = dy)) +
  coord_equal() +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 6))
```

### Cell count inconsistency near the origin

The asymmetry has a concrete consequence when data straddles the global
origin — which the North Atlantic, the main theatre of EU VMS activity,
conspicuously does. This was first documented for the midpoint formula vs.
c-squares in a [2025 blog post](https://heima.hafro.is/~einarhj/fishydata/ramb/2025-06-18_on_gridding).

Consider 10,000 synthetic pings on a regular 0.001° grid spanning −0.05° to
+0.05° in both dimensions. This region contains exactly four 0.05° cells in a
symmetric 2×2 arrangement, so any sensible method should produce four cells
with 2,500 counts each:

```{r cell-counts}
dx <- 0.05
g_orig <- expand_grid(
  lon = seq(-0.05, 0.0499, by = 0.001),
  lat = seq(-0.05, 0.0499, by = 0.001)
) |>
  mutate(cs    = fd_calc_csq(lon, lat, dx),
         x_mid = lon %/% dx * dx + dx/2,
         y_mid = lat %/% dx * dx + dx/2)

csq_ll <- csq2lonlat(g_orig$cs, degrees = dx)
g_orig <- g_orig |>
  mutate(x_csq = csq_ll$lon,
         y_csq = csq_ll$lat)

cat("Midpoint — 4 balanced cells:\n")
g_orig |> count(lon = x_mid, lat = y_mid) |> as.data.frame()
```

```{r cell-counts-csq}
cat("C-square centroids — 9 cells:\n")
g_orig |> count(lon = x_csq, lat = y_csq) |> as.data.frame()
```

The midpoint formula delivers four balanced cells. C-squares produces
**nine**, scattering the 100 boundary pings into five additional fringe cells
centred at ±0.075°. The counts range from 1 to 2,500.

```{r cell-count-plot, fig.cap = "Ping counts per cell: midpoint (left) vs c-squares (right). The heavy lines mark the prime meridian and equator. Midpoint produces a symmetric 2×2 grid; c-squares fragments the same data into 9 cells."}
bind_rows(
  g_orig |> count(lon = x_mid, lat = y_mid) |> mutate(method = "midpoint"),
  g_orig |> count(lon = x_csq, lat = y_csq) |> mutate(method = "c-squares")
) |>
  filter(between(lon, -0.11, 0.11), between(lat, -0.11, 0.11)) |>
  ggplot(aes(lon, lat, fill = factor(n))) +
  theme_bw() +
  geom_tile(width = dx, height = dx, colour = "grey70") +
  scale_fill_brewer(palette = "Set1") +
  geom_hline(yintercept = 0, linewidth = 1.2) +
  geom_vline(xintercept = 0, linewidth = 1.2) +
  facet_wrap(~method) +
  labs(x = "Longitude", y = "Latitude", fill = "Count") +
  coord_equal()
```

To be clear: this is not a bug in `fd_calc_csq()`. It faithfully implements
the specification. The specification just happens to produce a result that
would surprise anyone expecting a standard raster.

## Three implementations, two conventions

If the specification were unambiguous, one would expect independent
implementations to agree. Consider the same two boundary positions as before:

```{r csquares-comparison}
as_csquares_lh <- function(lon, lat, dx) {
  matrix(c(lon, lat), ncol = 2) |>
    csquares::as_csquares(resolution = dx) |>
    pull(csquares) |>
    as.character()
}

data.frame(
  lon          = lons,
  lat          = lats,
  fd_calc_csq  = fd_calc_csq(lons, lats, 0.05),
  as_csquares  = as_csquares_lh(lons, lats, 0.05)
)
```

`fd_calc_csq()` and `vmstools::CSquare()` both implement the spec's
"away from origin" rule and produce identical results. `csquares::as_csquares()`
goes the other way at boundary points. Three independent implementations of a
14-page specification have arrived at two distinct conventions. This is,
at minimum, a sign that the format does not lend itself to straightforward
implementation.

For interior points all three agree, and GPS positions rarely fall exactly on
a 0.05° boundary, so the practical impact on any given dataset is small. The
point is not that maps will be noticeably wrong — it is that the format
introduces complexity and ambiguity in exchange for no analytical benefit.

## The performance penalty

C-squares encode coordinates by recursively extracting digits from the absolute
value of each coordinate and assembling them into a string. At 0.05° that is
a 14-character string assembled by four levels of `paste0()`. A typical
Icelandic fleet VMS dataset — around 1 million pings — gives a sense of the
cost:

```{r code-size}
set.seed(6271)
n <- 1e6
lon_r <- runif(n, -25, -13)
lat_r <- runif(n,  63,  67)

csq_vec <- fd_calc_csq(lon_r, lat_r, 0.05)
mid_lon <- lon_r %/% 0.05 * 0.05 + 0.025
mid_lat <- lat_r %/% 0.05 * 0.05 + 0.025

cat("C-square vector (1M rows):", format(lobstr::obj_size(csq_vec)), "\n")
cat("Two midpoint vectors:     ", format(lobstr::obj_size(mid_lon, mid_lat)), "\n")
```

R interns repeated strings efficiently, so memory is not the main concern
here. The practical issues are:

- **Encoding speed.** Multi-level integer arithmetic and `paste0()` versus a
  single vectorised arithmetic expression.
- **Grouping and joining.** String hash-tables are slower than numeric ones.
  For large aggregations — `count(csq)`, `group_by(csq) |> summarise(...)` —
  the difference is noticeable.
- **DuckDB and Arrow incompatibility.** The midpoint formula translates
  directly to SQL arithmetic and works on any lazy table. `fd_calc_csq()` is
  an R function and cannot be called inside a DuckDB query.
- **Raster interoperability.** Midpoint coordinates slot directly into
  `stars`, `terra`, and any raster pipeline. C-square strings require a
  decode step.

Putting numbers to the first point:

```{r encode-bench}
library(bench)
library(vmstools)

d_bench <- tibble(lon = lon_r, lat = lat_r)

enc_bench <- bench::mark(
  midpoint         = d_bench |>
    mutate(x = lon %/% dx * dx + dx/2, y = lat %/% dx * dx + dx/2),
  fd_calc_csq      = d_bench |>
    mutate(cs = fd_calc_csq(lon, lat, dx)),
  vmstools_CSquare = d_bench |>
    mutate(cs = vmstools::CSquare(lon = lon, lat = lat, degrees = dx)),
  check = FALSE,
  min_iterations = 3L
)
data.frame(
  expression = as.character(enc_bench$expression),
  min        = format(enc_bench$min),
  median     = format(enc_bench$median),
  mem_alloc  = format(enc_bench$mem_alloc)
)
```

Roughly 200 times slower than the midpoint formula, and 500 times slower than
that for `vmstools::CSquare()`. Memory allocation follows a similar pattern.
Two hundred times slower is, by most standards, a fairly compelling argument.

The aggregation picture is similar. When the goal is a gridded summary — count,
catch, or effort per cell — DuckDB can push both encoding and the group-by
into a single in-memory query. For c-squares there is no such path:

```{r agg-bench}
library(duckdb)

con <- dbConnect(duckdb())
duckdb_register(con, "pings", d_bench)
q <- tbl(con, "pings")

agg_bench <- bench::mark(
  r_midpoint  = d_bench |>
    mutate(x = lon %/% dx * dx + dx/2, y = lat %/% dx * dx + dx/2) |>
    count(x, y),
  r_csq       = d_bench |>
    mutate(cs = fd_calc_csq(lon, lat, dx)) |>
    count(cs),
  db_midpoint = q |>
    mutate(x = lon %/% dx * dx + dx/2, y = lat %/% dx * dx + dx/2) |>
    count(x, y) |>
    collect(),
  check = FALSE,
  min_iterations = 3L
)
data.frame(
  expression = as.character(agg_bench$expression),
  min        = format(agg_bench$min),
  median     = format(agg_bench$median),
  mem_alloc  = format(agg_bench$mem_alloc)
)
dbDisconnect(con)
```

DuckDB encodes and aggregates 1 million pings to a gridded count while
allocating almost no R memory — the computation stays inside the DuckDB
process and only the result table comes back. The c-square workflow has no
equivalent; `r_csq` is the only option available.

### The coming data volume problem

The EU VMS regulation currently mandates a minimum ping frequency of two hours.
At that rate, a fleet of 1,000 vessels generates roughly 12,000 pings per day —
a volume that most modern hardware handles without breaking a sweat, even with
the c-square overhead.

This will not remain the situation for long. AIS data — already collected by
coastal states and increasingly integrated into fisheries monitoring — operates
on a fundamentally different scale. AIS transponders typically transmit every
few seconds when under way; even aggregated to one-minute resolution, a fleet
of 1,000 vessels generates around 1.4 million pings per day, roughly 120 times
the VMS volume. Aggregated to national or pan-European scale, annual AIS
datasets run into the billions of rows.

At that scale the 200× encoding penalty becomes genuinely prohibitive. DuckDB
and Arrow are not optional extras in a billion-row workflow — they are
necessities. And as shown above, c-squares do not work inside either.

The data call will, sooner or later, need to accommodate AIS. When it does, the
question of whether to continue using a metadata indexing format designed for
catalogue search — at a resolution 400 times finer than its own spec recommends,
processed at 200 times the cost of the alternative — will answer itself.

## The alternative: one line of arithmetic

For any fixed resolution `dx`, the midpoint of the cell containing coordinate
`x` is:

```r
x_mid <- x %/% dx * dx + dx / 2
```

The `%/%` operator truncates toward negative infinity (floor semantics),
giving a **left-closed, right-open** interval at every coordinate regardless
of sign:

```{r midpoint-demo}
dx <- 0.05
lons_test <- c(-0.10, -0.09, -0.05, -0.049999, 0.0, 0.049999, 0.05, 0.10)

data.frame(
  lon      = lons_test,
  midpoint = lons_test %/% dx * dx + dx / 2,
  csq      = fd_calc_csq(lons_test, rep(0.025, length(lons_test)), dx)
)
```

`lon = −0.05` and `lon = −0.049999` share the same midpoint; `fd_calc_csq()`
assigns them to different cells. The formula requires no string construction,
no hierarchical encoding, no lookup table, and no external package. It works
identically in R, dplyr, Arrow, and DuckDB. It produces a symmetric raster at
every coordinate.

The one thing it cannot do is produce a c-square code. For the intermediate
steps of a VMS workflow — binning pings, aggregating catch, computing swept
area, joining to raster habitat layers — this is not a limitation. For the
final submission to ICES, it currently is. The appropriate response is to
generate c-square codes at the very last step, from the already-aggregated
result table, where the number of rows is a few thousand grid cells rather
than millions of individual pings, and the encoding cost is negligible.

## Summary

| | `fd_calc_csq` | `vmstools::CSquare` | `csquares::as_csquares` | midpoint |
|---|:---:|:---:|:---:|:---:|
| Spec-compliant boundary | ✓ | ✓ | ✗ | n/a |
| Consistent with standard raster | ✗ | ✗ | ✗ | ✓ |
| DuckDB / Arrow compatible | ✗ | ✗ | ✗ | ✓ |
| Multi-resolution nesting | ✓ | ✓ | ✓ | ✗ |
| No boundary ambiguity | ✗ | ✗ | ✗ | ✓ |
| Relative encoding speed | ~200× | ~500× | — | 1× |
| Viable at AIS frequency | ✗ | ✗ | ✗ | ✓ |

The recommendation is to use the midpoint formula throughout the analytical
pipeline and reserve c-square encoding for the submission output step, where
it adds negligible cost.

If the data call were being designed today, with AIS on the horizon and DuckDB
on the desktop, one suspects the format choice would be different.

## References

Rees, T. (2003). "C-squares, a new spatial indexing system and its
applicability to the description of oceanographic datasets." *Oceanography*,
16(1), 11–19. <https://www.cmar.csiro.au/csquares/spec1-1.htm>
